---
title: "Master_Data_Prep"
author: "Conor Loy"
date: "June 5, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### This R Markdown file  creates a miRNA expression file from data queried from the GDC API. The following code averages expression for individual miRNA from different tissue samples seperated by tumor or nomal status. The data used to create these expression means is miRNA-seq cpm (counts per million) values. To minimize flooding the global enviroment lists of dataframes will be used throughout this file. Also, this allows for reproducability if another tissue sample would like to be added.

## Load Packages and Read in Data


##### First, all packages and data must be read loaded. Data containing miRNA-seq cpm expression values queried from the GDC will be used. The package dplyr is required for data wrangling.



```{r,echo=TRUE,warning=FALSE}

library(knitr)
library(rmarkdown)
library(dplyr)


BLCA_cpm <- read.csv("../../CleanData/BLCA_cpm.csv")

BRCA_cpm <- read.csv("../../CleanData/BRCA_cpm.csv")

COAD_cpm <- read.csv("../../CleanData/COAD_cpm.csv")

PRAD_cpm <- read.csv("../../CleanData/PRAD_cpm.csv")
```


## Seperate Tumor and Normal Tissue Samples


##### The following code chunk contains lists that will be used to create dataframes for normal and tumor tissue samples.


```{r}
orig_df <- list(BLCA_cpm,BRCA_cpm,COAD_cpm,PRAD_cpm)
source <- c("blca","brca","coad","prad")
```


##### The following code creates a list object 'df_list_tum' that contains filtered dataframes of the orinial data with only tumor samples. 


```{r}
df_list_tum <- sapply(orig_df, function(x) filter(x,Site=="Tumor"), simplify=FALSE)          # Create a list with four filtered dataframes in it

df_list_tum <- setNames(df_list_tum, paste0(source, "_tum"))                                 # Name the dataframes to indicate tumor status


# Test
blca_tum <- BLCA_cpm %>%
  filter(Site == "Tumor")

all.equal(df_list_tum$blca_tum,blca_tum)

rm(blca_tum)
```
  
  
##### The following code creates a list object 'df_list_nrm' that contains filtered dataframes of the orinial data with only normal samples.


```{r}
df_list_nrm <- sapply(orig_df, function(x) filter(x,Site=="Normal"), simplify=FALSE)          # Create a list with four filtered dataframes in it

df_list_nrm <- setNames(df_list_nrm, paste0(source, "_nrm"))                                  # Name the dataframes to indicate normal status

blca_nrm <- BLCA_cpm %>%
  filter(Site == "Normal")


all.equal(df_list_nrm$blca_nrm,blca_nrm)

rm(blca_nrm)                                                                                  
```


## Average miRNA Expression


##### The following code chunk contains lists that will be used to create lists of dataframes for normal and tumor tissue sample miRNA expression.


```{r}
df_source <- c("BLCA","BRCA","COAD","PRAD")
miRNA_names <- colnames(BLCA_cpm[,!names(BLCA_cpm) %in% c("UUID","Site")])
```

### Tumor Samples

##### The following code creates a list of dataframes containing average expression values for miRNA in specific tumor tissue samples. This is done by first creating a list of empty dataframes for each tissue type, then populating each data frame one row at a time with average miRNA expression values.

```{r}
tum_df <- sapply(df_source, function(x) data.frame(miRNA=NA,tumor=NA), simplify=FALSE)        # Create list of empty dataframes    

tum_df <- setNames(tum_df, paste0(df_source, "_sumT"))                                        # Properly name dataframes in list

column_Names <- c("BLCA_Tumor","BRCA_Tumor","COAD_Tumor","PRAD_Tumor")                        # Column names to distinguish columns in master file


for (i in 1:length(tum_df)){
  
  temp <- df_list_tum[[i]]                                                                    # Use dataframe in list of filtered dataframes from earlier
  
  for (miRNA in miRNA_names){                                                                 # Iterate over miRNA names
    
    exp <- c(miRNA,mean(temp[[miRNA]]))                                                       # Find mean expression of indivual miRNA
    
    tum_df[[i]]<-rbind(tum_df[[i]],exp)                                                       # Add to dataframe being created
  }

  colnames(tum_df[[i]])[2] = column_Names[i]  

    }

merged_tum <- Reduce(function(x,y) merge(x,y,by="miRNA"),tum_df)                              # Merge all tunmor summary tables together

```


### Normal Samples

##### The following code creates a list of dataframes containing average expression values for miRNA in specific normal tissue samples. This is done by first creating a list of empty dataframes for each tissue type, then populating each data frame one row at a time with average miRNA expression values. The code is identical to the code above apart from tissue sample type.
```{r}
nrm_df <- sapply(df_source, function(x) data.frame(miRNA=NA,normal=NA), simplify=FALSE)

nrm_df <- setNames(nrm_df, paste0(df_source, "_sumN"))

column_Names <- c("BLCA_Normal","BRCA_Normal","COAD_Normal","PRAD_Normal")


for (i in 1:length(nrm_df)){
  
  temp <- df_list_nrm[[i]]
  
  for (miRNA in miRNA_names){
    
    exp <- c(miRNA,mean(temp[[miRNA]]))
    
    nrm_df[[i]]<-rbind(nrm_df[[i]],exp)
  }

  colnames(nrm_df[[i]])[2] = column_Names[i]  

    }



merged_nrm <- Reduce(function(x,y) merge(x,y,by="miRNA"),nrm_df)

```


## Merge, Format and Download Master File

##### The normal and tumor summary tables are merged and the columns are ordered in a logical manner.

```{r}

final <- merge(merged_nrm, merged_tum,by="miRNA")

final <- final[c("miRNA",	"COAD_Normal",	"COAD_Tumor",	"BRCA_Normal",	"BRCA_Tumor",	"BLCA_Normal",	"BLCA_Tumor",	"PRAD_Normal",	"PRAD_Tumor")]
  
#write.csv(final, "../../CleanData/expression.csv")
```

